Linear vertical shear instability in astrophysical disks: a semi-numerical study

### 2014-12-23 linear_iso.tex
replace numerical section with that of full equations
added nearly isothermal discussion, only consider special case of fundamental mode for now. should generalize

\section{Analytical discussion of vertically isothermal disks} 
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the shearing box. We
examine separately the limiting case of isothermal perturbations
($\gamma\equiv1$) and adiabatic perturbations ($\gamma\neq 1$ and
$t_c\to\infty$).   
\subsection{Isothermal perturbations}
The linear problem simplifies considerably for the case
$\Gamma=\gamma=1$. Then 
\begin{align}
  Q=W. 
\end{align}
Inserting this into Eq. \ref{ode_w} gives a single equation for the linear stability
problem for isothermal perturbations in a vertically isothermal disk, 
\begin{align}\label{iso_ode}
  \frac{d^2W}{dz^2} + \left(\frac{d\ln{\rho}}{dz} - \frac{\ii k_x
      r}{D}\frac{d\Omega^2}{dz}\right) \frac{dW}{dz} +
  \sigma^2\left(\frac{1}{c_s^2} + \frac{k_x^2}{D}\right)W=0, 
\end{align} 
subject to appropriate boundary conditions. We remark that Eq
.\ref{iso_ode} is also applicable for    
$\gamma=\Gamma\neq 1$ in the limit $t_c\to\infty$ (i.e. neutrally
stratified adiabatic disks), since in that case  
$Q\simeq W$ as well. However, the following discussion is only
valid for $\gamma=\Gamma=1$ because $c_s$ is taken to be a constant.

@@

\subsubsection{Polynomial solutions with vertical shear}\label{iso_poly}
We can seek power-series solution to Eq. \ref{iso_ode3},
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients must satisfy the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[n - l\left(1+\ii \epsilon q  \hat{k}\right)\right] a_l = 0, 
\end{align}
where $n$ is not neccessarily an integer. Indeed, if we demand
a polynomial of degree $L$ as the solution, then the eigenfrequency
must satisfy
\begin{align}
\hat{\sigma}^2 = L\left(\frac{1+\ii q \epsilon
    \hat{k}}{1+\hat{k}^2}\right).
\end{align}
The eigenfrequency $\hat{\sigma}$ is therefore complex for
$L\geq1$. We are interested in 
unstable modes for which $\imag{\hat{\sigma}}>0$ is the growth rate and is
given by 
\begin{align}\label{simple_growth}
  \hat{\nu} =\sqrt{
   \frac{L}{2\left(1+\hat{k}^2\right)}\left(\sqrt{1+q^2\epsilon^2\hat{k}^2} - 
    1\right)}. 
\end{align}
For fixed $q$ and $\epsilon$, the growth rate vanishes for both
$\hat{k}^2\to0$ and $\hat{k}^2\to\infty$. The maximum growth rate
occurs at the optimum wavenumber $\hat{k}_\mathrm{opt}$
\begin{align}
  |\hat{k}_\mathrm{opt}| = \sqrt{\frac{2+|\epsilon q|}{|\epsilon q|}},
\end{align}
and
\begin{align}
  \mathrm{max}\left(\hat{\nu}\right) =\frac{\sqrt{L}|\epsilon
    q|}{2\sqrt{1+|\epsilon q|}}. \label{iso_max_growth}
\end{align}
Since $L\geq1$, there is no limit to the maximum growth rate. This is
an artifact of the thin-disk approximation because the vertical 
shear $d\Omega^2/dz\propto z$ increases indefinitely with
height. However, we find Eq. \ref{iso_max_growth} compares well with
numerical solutions of the governing equation for small $L$.

@@

>>>>

\section{Analytical discussion of vertically isothermal disks} 
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the shearing box. We
examine separately the limiting case of isothermal perturbations
($\gamma\equiv1$ and any value of $t_c$), nearly-isothermal
perturbations ($\gamma\geq 1$ and $t_c\to0$), and 
adiabatic perturbations ($\gamma\neq 1$ and $t_c\to\infty$).   
\subsection{Isothermal perturbations}\label{iso_discuss}
The linear problem simplifies considerably for the case
$\Gamma=\gamma=1$.  Then 
\begin{align}
  Q=W. 
\end{align}
Inserting this into Eq. \ref{ode_w} gives a single equation for the linear stability
problem for isothermal perturbations in a vertically isothermal disk, 
\begin{align}\label{iso_ode}
  \frac{d^2W}{dz^2} + \left(\frac{d\ln{\rho}}{dz} - \frac{\ii k_x
      r}{D}\frac{d\Omega^2}{dz}\right) \frac{dW}{dz} +
  \sigma^2\left(\frac{1}{c_s^2} + \frac{k_x^2}{D}\right)W=0, 
\end{align} 
subject to appropriate boundary conditions. Eq. \ref{iso_ode} also
applies to $\gamma>1$ for $t_c\to0$.

We remark that Eq.\ref{iso_ode} is also applicable for    
$\gamma=\Gamma\neq 1$ in the limit $t_c\to\infty$ (i.e. neutrally
stratified adiabatic disks), since in that case  
$Q\simeq W$ as well. However, the following discussion is only
valid for $\gamma=\Gamma=1$ because $c_s$ is taken to be a constant.

@@

\subsubsection{Polynomial solutions with vertical shear}\label{iso_poly}
We can seek power-series solution to Eq. \ref{iso_ode3},
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients must satisfy the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[n - l\left(1+\ii \epsilon q  \hat{k}\right)\right] a_l = 0, 
\end{align}
where $n$ is not neccessarily an integer. Indeed, if we demand
a polynomial of degree $L$ as the solution, then the eigenfrequency
must satisfy
\begin{align}\label{sig2_iso}
\hat{\sigma}^2 = L\left(\frac{1+\ii q \epsilon
    \hat{k}}{1+\hat{k}^2}\right).
\end{align}
The eigenfrequency $\hat{\sigma}$ is therefore complex for
$L\geq1$. We are interested in 
unstable modes for which $\imag{\hat{\sigma}}>0$ is the growth rate and is
given by 
\begin{align}\label{simple_growth}
  \hat{\nu} =\sqrt{
   \frac{L}{2\left(1+\hat{k}^2\right)}\left(\sqrt{1+q^2\epsilon^2\hat{k}^2} - 
    1\right)}. 
\end{align}
For fixed $q$ and $\epsilon$, the growth rate vanishes for both
$\hat{k}^2\to0$ and $\hat{k}^2\to\infty$. The maximum growth rate
occurs at the optimum wavenumber $\hat{k}_\mathrm{opt}$
\begin{align}
  |\hat{k}_\mathrm{opt}| = \sqrt{\frac{2+|\epsilon q|}{|\epsilon q|}},
\end{align}
and
\begin{align}
  \mathrm{max}\left(\hat{\nu}\right) =\frac{\sqrt{L}|\epsilon
    q|}{2\sqrt{1+|\epsilon q|}}. \label{iso_max_growth}
\end{align}
Since $L\geq1$, there is no limit to the maximum growth rate. This is
an artifact of the thin-disk approximation because the vertical 
shear $d\Omega^2/dz\propto z$ increases indefinitely with
height. However, we find Eq. \ref{iso_max_growth} compares well with
numerical solutions of the governing equation for small $L$.

@@

\subsection{Nearly-isothermal perturbations}
We can use the explicit solution developed above to examine the effect
of cooling $t_c$ on perturbations with $\gamma>1$. We shall consider
the limit $t_c\to0$, for which perturbations can be considered
nearly-isothermal.

We begin by writing the pair of ODEs for $W$ and $Q$
(Eq. \ref{ode_w}---\ref{ode_Q}) in the thin-disk, low-frequency, 
nearly-Keplerian approximations as used previously, as well as
adopting dimensionless co-ordinates and variables, 
\begin{align}
  &\hat{\sigma}^2\left(Q +  \hat{k}^2W\right) \notag\\ 
  &= \hat{z} \left(1 +
    \ii\epsilon q \hat{k} \right)\left[W^\prime - \zhat \left(W -
      Q\right)\right] - \left[W^\prime - \zhat \left(W -
      Q\right)\right]^\prime\label{nearly_iso1},\\
&\beta \hat{\sigma}^2\left(W - \gamma Q\right) + \ii\hat{\sigma}
    \left(W-Q\right)\notag\\
   & = -\beta\zhat\left(\gamma-1\right) \left[W^\prime - \zhat \left(W -
      Q\right)\right]\label{nearly_iso2},
\end{align}
where we recall $\beta = t_c\Omega_k$ is the dimensionless cooling
time. It is evident that for $\beta\equiv 0$, we recover
Eq. \ref{iso_ode3}.

Since we have developed solutions for $\beta\equiv 0$, we may ask how
do the eigenfrequencies and eigenfunctions change when we change
$\beta$ from zero to a small but finite value? For sufficiently small
$\beta$, we expect the new solutions to only differ slighly from that
above. Then we may linearize Eq. \ref{nearly_iso1}---\ref{nearly_iso2} to
determine the effect of adding small finite cooling.

Let 
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, W\to W+\delta W,\, Q \to Q+\delta
  Q,\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align} 
where $W=Q$ is the polynomial solution described in \S\ref{iso_poly}
and $\hat{\sigma}^2$ is given by Eq. \ref{sig2_iso}. Inserting
Eq. \ref{nearly_iso_pert} into
Eq. \ref{nearly_iso1}---\ref{nearly_iso2} and keeping only first order
terms, we obtain
\begin{align}
  &2\hat{\sigma}\delta\hat{\sigma}\left(1+\hat{k}^2\right)W +
  \hat{\sigma}^2 \left(\delta Q + \hat{k}^2\delta W\right)\notag\\
  &= \zhat \left(1+\ii\epsilon q \hat{k}\right)\left[\delta W^\prime -
    \zhat \left(\delta W - 
      \delta Q\right)\right] - \delta W^{\prime\prime}\notag\\
  &\phantom{=} + \left[\zhat \left(\delta W -
     \delta Q\right)\right]^\prime,\label{nearly_iso3}\\
&\ii\hat{\sigma} \left(\delta W - \delta Q\right) =
\delta\beta\left(\gamma-1\right)\left(\hat{\sigma}^2W - \hat{z}W^\prime\right).\label{nearly_iso4}
\end{align}

We could proceed further by considering the set of polynomial
solutions for $W$ described in \S\ref{iso_poly} altogether, but it is
simplest to look at the fundamental mode. Thus
we consider the `basic state'
\begin{align*}
  W = \hat{z},\quad \hat{\sigma}^2\left(1+\hat{k}^2\right) = 1 +
  \ii\epsilon q \hat{k},
\end{align*}
then we can assume
\begin{align}
  \delta W = b \zhat^3,
\end{align}
where $b$ is a constant to be determined. Eliminating $\delta Q$ from
Eq. \ref{nearly_iso3}---\ref{nearly_iso4}, inserting the above
expressions for $W$ and $\delta W$ and balancing coefficients of
$\zhat$ and $\zhat^3$, we find
\begin{align}
 & 2\left(1+\ii\epsilon q \hat{k}\right) \delta\hat{\sigma} \notag\\& = -6b\hat{\sigma} -
2\ii \delta \beta\left(\gamma-1\right)\left(\hat{\sigma}^2-1\right) - \ii
  \hat{\sigma}^2\delta\beta\left(\gamma-1\right),\\
& 0=  2b\hat{\sigma}+\ii\delta\beta
\left(\gamma-1\right)\left(\hat{\sigma}^2 - 1\right),
\end{align}
which gives
\begin{align}\label{nearly_iso_dsig}
  \delta \hat{\sigma} =
  -\frac{\ii\left(\gamma-1\right)\hat{k}^2\left(\ii\epsilon
    q - \hat{k}\right)^2}{2\left(1+\ii\epsilon q \hat{k}\right)\left(1+\hat{k}^2\right)^2}\delta\beta.
\end{align}
The imaginary part of Eq. \ref{nearly_iso_dsig} is
\begin{align}
   \delta\hat{\nu} =
  -\frac{\left(\gamma-1\right)\hat{k}^2 \left(\hat{k}^2 -
      2\epsilon^2q^2\hat{k}^2 - \epsilon^2q^2\right)}{2\left(1+\epsilon^2 q^2
        \hat{k}^2\right)\left(1+\hat{k}^2\right)^2}\delta\beta.  
\end{align}
Since $\epsilon \ll 1$, introducing finite cooling $\delta\beta>0$
implies $\delta\nu < 0$, i.e. stabilization. Interestingly, for both
$\hat{k}\to0$ and $\hat{k}\to\infty$ we find $\delta\hat{\nu}\to0$, so
stabilization by finite cooling is ineffective at very large or very small scales.
